Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2024-12-09 _tex/index.tex
Built site for gh-pages

Figure~\ref{fig-discriminal_process} illustrates the discriminal
processes along a quality trait continuum for two written texts. The
figure shows that these processes individually follow a Normal
distribution. Furthermore, it shows that the modal discriminal process
for Text B is positioned further along the continuum than Text A
\((S_{B} > S_{A})\). Additionally, the figure highlights the broader
distribution of Text B compared to Text A as a result of its larger
discriminal dispersion \((\sigma_{B} > \sigma_{A})\).

@@

Figure~\ref{fig-discriminal_difference} illustrates the distribution of
the discriminal difference for the texts presented in
Figure~\ref{fig-discriminal_process}. The figure indicates that the
judge perceives Text B as having significantly higher quality than Text
A. This conclusion is evident from the positive difference in the modal
discriminal processes of the texts \((S_{B} - S_{A} > 0)\) and the area
under the curve where the discriminal difference distinctly favors Text
B over Text A, denoted as \(P(B > A)\) (shaded gray area). As a result,
the dichotomous outcome of this comparison is more likely to favor Text
B over Text A.

@@

The previous section outlines the general form of Thurstone's theory,
which applies to a CJ design where a single judge evaluates multiple
stimuli. For the practical application of the theory, Thurstone
developed four additional cases derived from this general form, where
each successive case incorporates additional simplifying assumptions.
Case I represents the general form of the theory. Case II extends this
by allowing multiple judges to make comparisons rather than restricting
the comparisons to a single judge. Case III introduces the assumption of
zero correlation between stimuli. Case IV builds on this by assuming the
stimuli have similar dispersions. Finally, Case V replaces this
assumption with the condition that the stimuli have equal discriminal
dispersions. Table~\ref{tbl-thurstone_cases} summarizes these cases and
their assumptions. For a detailed discussion of this progression, refer
to \citet{Thurstone_1927b} and \citet[pp.~248-253]{Bramley_2008}.

@@

Nevertheless, Thurstone originally developed Case V to provide a
``rather coarse scaling'' of traits \citep[pp.~269]{Thurstone_1927b},
prioritizing statistical simplicity over precision in trait measurement
\citep[pp.~677]{Kelly_et_al_2022}. He explicitly warned against its
untested application, stating that its use ``should not be made without
(an) experimental test'' \citep[pp.~270]{Thurstone_1927b}, acknowledging
that some assumptions might be problematic when complex traits or
heterogeneous stimuli are involved \citep[pp.~376]{Thurstone_1927a}.
Consequently, given that modern CJ applications frequently involve such
traits and stimuli, two main assumptions of Case V and, by extension, of
the BTL model may not consistently hold in theory or practice: the
assumption of equal dispersion and zero correlation between stimuli.

@@

The discriminal dispersions of stimuli are crucial in determining the
dichotomous outcomes of pairwise comparisons. Discrepancies in these
dispersions shape the distribution of the discriminal difference,
directly influencing the comparison outcome. Figure~\ref{fig-dispersion}
illustrates how more uncertainty in the trait perception of one text
relative to another broadens the distribution of their discriminal
difference. This broadening affects the area under the curve where the
discriminal difference distinctly favors one text over the other,
\(P(B > A)\). Additionally, the figure shows that when the discriminal
dispersions of the texts are equal \((\sigma_{B}-\sigma_{A}=0)\), the
discriminal difference is more likely to favor one text over the other
than when their dispersions differ (shaded gray area).

@@

\caption{\label{fig-dispersion}The Discrepancy in the dispersions of
stimuli and their effect on the distribution of the discriminal
difference}

@@

However, Thurstone contended that the assumption of equal dispersions
may not hold when researchers assess complex traits or heterogeneous
stimuli \citep[pp.~376]{Thurstone_1927a}, as these traits and stimuli
can introduce judgment discrepancies due to their unique features
\citep{vanDaal_et_al_2016, Lesterhuis_2018, Chambers_et_al_2022}.
Indeed, evidence of such violation may already exist in the CJ
literature as misfit statistics. These statistics measure the judgment
discrepancies associated with a given stimulus
\citetext{\citealp[pp.~12]{Pollitt_2004}; \citealp[pp.~20]{Goossens_et_al_2018}}.
For instance, labeling texts as ``misfits'' indicates that comparisons
involving these texts result in more judgment discrepancies than others
\citep{Pollitt_2012a, Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018}.
This finding implies that discriminal differences associated with
``misfits'' texts usually display a broader dispersion, suggesting that
their discriminal processes also exhibit more variation than other
texts. Notably, this reasoning also applies to ``misfit'' judges, whose
evaluations reflect substantial deviations from the shared consensus due
to the unique characteristics of the stimuli or the judges themselves.
Moreover, the presence of these ``misfit'' judges and their deviations
can introduce additional statistical and measurement issues, as
discussed in Section~\ref{sec-theory-issue1b}.

Therefore, assuming equal dispersions between stimuli, despite its
violation, can cause Case V (and the BTL model) to inflate the
reliability of the comparison outcome. Moreover, ignoring the difference
in dispersions can lead to the neglect of critical differences in the
reliability of the trait across stimuli, resulting in inaccurate
conclusions about the trait's estimates \citep[pp.~370]{McElreath_2020}.
Furthermore, if researchers acknowledge that misfit statistics help
identify these critical differences, the usual practice in the CJ
literature of excluding stimuli based on these statistics
\citep{Pollitt_2012b, vanDaal_et_al_2017, Goossens_et_al_2018} risks
discarding valuable information and introducing bias into the trait's
estimates \citep[chap.~12]{Zimmerman_1994, McElreath_2020}. The
direction and magnitude of these biases are unpredictable, as they
depend on the specific stimuli researchers exclude from the analysis.
Together, these oversights undermine the reliability of the trait and
ultimately compromise its validity \citep[pp.~2]{Perron_et_al_2015}.

@@

Similar to the discriminal dispersions, the correlation between
discriminal processes plays a crucial role in determining the
dichotomous outcomes of pairwise comparisons. This correlation also
shapes the distribution of the discriminal difference, directly
influencing the comparison outcome. Figure~\ref{fig-correlation}
illustrates how the dependence in trait perception between two texts
narrows the distribution of their discriminal difference. This narrowing
affects the area under the curve where the discriminal difference
distinctly favors one text over the other, \(P(B>A)\). Moreover, the
figure shows that when two texts are independent or uncorrelated
\((\rho=0)\), their discriminal difference is less likely to favor one
text over the other than when the texts are highly correlated (shaded
gray area).

@@

In the second scenario, the shared context or inherent connections
created by additional hierarchical structures may further introduce
dependencies between stimuli, a statistical phenomenon commonly known as
clustering \citep{Everitt_et_al_2010}. Although the CJ literature
acknowledges the presence of such hierarchical structures, the
statistical handling of this extra source of dependency between stimuli
has been inadequate. For example, when CJ data includes multiple samples
of stimuli from the same individuals, researchers often rely on
(average) estimated BTL scores to conduct subsequent analyses and tests
at the individual hierarchical level
\citep{Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.
This approach, however, can introduce additional statistical and
measurement issues, as discussed in Section~\ref{sec-theory-issue2}.

In any case, the psychometric and statistical literature emphasizes the
need to address factors that create dependencies between stimuli, as
failing to do so can affect the reliability of the comparison outcomes
and lead to inaccurate conclusions about the trait's estimates. For
instance, researchers who overlook additional relevant traits, such as
judges' biases, can cause dimensional mismatches in the statistical
model used for analysis. This mismatch can artificially inflate the
reliability of the trait \citep[pp.~341]{Hoyle_et_al_2023} or, even
worse, introduce bias into the trait's estimates \citep{Ackerman_1989}.
Furthermore, as discussed in Section~\ref{sec-theory-issue1a},
researchers who exclude judges based on misfit statistics can risk
discarding valuable information, further biasing the trait's estimates
\citep[chap.~12]{Zimmerman_1994, McElreath_2020}. Lastly, researchers
who fail to account for hierarchical (grouping) structures can reduce
the precision of model parameter estimates, which may amplify the
overestimation of reliability \citep[pp.~482]{Hoyle_et_al_2023}.
Together, these issues undermine the reliability of the trait and
compromise the validity of its estimates
\citep[pp.~2]{Perron_et_al_2015}.

@@

@@

>>>>

Figure~\ref{fig-discriminal_process} illustrates the discriminal
processes along a quality trait continuum for two written texts. The
figure shows that each discriminal process follows a Normal
distribution. It also indicates that the modal discriminal process for
Text B is positioned further along the continuum than that of Text A
\((S_{B} > S_{A})\), signifying a higher quality for Text B.
Additionally, the figure highlights that Text B has a broader
distribution compared to Text A, which arises from its larger
discriminal dispersion \((\sigma_{B} > \sigma_{A})\).

@@

Figure~\ref{fig-discriminal_difference} illustrates the distribution of
the discriminal difference for the texts presented in
Figure~\ref{fig-discriminal_process}. The figure indicates that the
judge perceives Text B as having significantly higher quality than Text
A. This conclusion is evident from the positive difference in their
modal discriminal processes \((S_{B} - S_{A} > 0)\) and the area under
the curve where the discriminal difference distinctly favors Text B over
Text A, denoted as \(P(B > A)\) (shaded gray area). As a result, the
dichotomous outcome of this comparison is more likely to favor Text B
over Text A.

@@

Since the general form of Thurstone's theory, outlined in
Section~\ref{sec-thurstone_theory}, applies to a CJ design where a
single judge evaluates multiple stimuli, Thurstone developed four
additional cases for the theory's practical application, each
progressively incorporating additional simplifying assumptions. Case I
represents the general form of the theory. Case II extends this by
allowing multiple judges to make comparisons rather than restricting the
comparisons to a single judge. Case III introduces the assumption of
zero correlation between stimuli. Case IV builds on this by assuming the
stimuli have similar dispersions. Finally, Case V replaces this
assumption with the condition that the stimuli have equal discriminal
dispersions. Table~\ref{tbl-thurstone_cases} summarizes these cases and
their assumptions. For a detailed discussion of this progression, refer
to \citet{Thurstone_1927b} and \citet[pp.~248-253]{Bramley_2008}.

@@

Nevertheless, Thurstone originally developed Case V to provide a
``rather coarse scaling'' of traits \citep[pp.~269]{Thurstone_1927b},
prioritizing statistical simplicity over precision in trait measurement
\citep[pp.~677]{Kelly_et_al_2022}. Thurstone explicitly warned against
its untested application, stating that its use ``should not be made
without (an) experimental test'' \citep[pp.~270]{Thurstone_1927b},
acknowledging that some assumptions could be problematic when
researchers asesss complex traits or heterogeneous stimuli
\citep[pp.~376]{Thurstone_1927a}. Consequently, given that modern CJ
applications frequently involve such traits and stimuli, two main
assumptions of Case V and, by extension, of the BTL model may not
consistently hold in theory or practice: the assumption of equal
dispersion and zero correlation between stimuli.

@@

The discriminal dispersions of stimuli are crucial in determining the
dichotomous outcomes of pairwise comparisons. Holding all other modeling
factors constant, discrepancies in these dispersions shape the
distribution of the discriminal difference, directly influencing the
comparison outcome. Figure~\ref{fig-dispersion} illustrates how more
uncertainty in the trait perception of one text relative to another,
\((\sigma_{B}-\sigma_{A})\), broadens the distribution of their
discriminal difference. This broadening affects the area under the curve
where the discriminal difference distinctly favors one text over the
other, \(P(B > A)\). Additionally, the figure shows that when the
discriminal dispersions of the texts are equal
\((\sigma_{B}-\sigma_{A}=0)\), the discriminal difference is more likely
to favor one text over the other than when their dispersions differ
(shaded gray area).

@@

\caption{\label{fig-dispersion}The discrepancy in the dispersions of
stimuli and their effect on the distribution of the discriminal
difference}

@@

However, Thurstone contended that this assumption of equal dispersions
may not hold when researchers assess complex traits or heterogeneous
stimuli \citep[pp.~376]{Thurstone_1927a}, as these traits and stimuli
can introduce judgment discrepancies due to their unique features
\citep{vanDaal_et_al_2016, Lesterhuis_2018, Chambers_et_al_2022}.
Indeed, evidence of such violation may already exist in the CJ
literature as misfit statistics. These statistics measure the judgment
discrepancies associated with a given stimulus
\citetext{\citealp[pp.~12]{Pollitt_2004}; \citealp[pp.~20]{Goossens_et_al_2018}}.
For instance, labeling texts as ``misfits'' indicates that comparisons
involving these texts result in more judgment discrepancies than others
\citep{Pollitt_2012a, Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018}.
This finding implies that discriminal differences associated with
``misfits'' texts usually display a broader dispersion, suggesting that
the discriminal processes of these texts also exhibit more variation
than other texts. Notably, this reasoning also applies to ``misfit''
judges, whose evaluations reflect substantial deviations from the shared
consensus due to the unique characteristics of the stimuli or the judges
themselves. Moreover, the presence of these ``misfit'' judges and their
deviations can introduce additional statistical and measurement issues,
which we discuss in Section~\ref{sec-theory-issue1b}.

Therefore, assuming equal dispersions between stimuli, despite its
violation, can cause Case V (and the BTL model) to inflate the
reliability of the outcome, resulting in inaccurate conclusions about
the comparison. Moreover, ignoring the disparity in dispersions can lead
to the neglect of critical differences in the reliability of the trait
across these stimuli, leading to erroneous conclusions about the trait's
estimates \citep[pp.~370]{McElreath_2020}. Additionally, if researchers
acknowledge that misfit statistics help identify these critical
differences in dispersion, the usual practice in the CJ literature of
excluding stimuli based on these statistics
\citep{Pollitt_2012b, vanDaal_et_al_2017, Goossens_et_al_2018} risks
discarding valuable information and introducing bias into the trait's
estimates \citep[chap.~12]{Zimmerman_1994, McElreath_2020}. The
direction and magnitude of these biases are often unpredictable, as they
depend on the specific stimuli researchers exclude from the analysis.

@@

Similar to the discriminal dispersions, the correlation between
discriminal processes plays a crucial role in determining the
dichotomous outcomes of pairwise comparisons. Holding all other modeling
factors constant, this correlation shapes the distribution of the
discriminal difference, directly influencing the comparison outcome.
Figure~\ref{fig-correlation} illustrates how the dependence in trait
perception between two texts narrows the distribution of their
discriminal difference. This narrowing affects the area under the curve
where the discriminal difference distinctly favors one text over the
other, \(P(B>A)\). Moreover, the figure shows that when two texts are
independent or uncorrelated \((\rho=0)\), their discriminal difference
is less likely to favor one text over the other than when the texts are
highly correlated (shaded gray area).

@@

In the second scenario, the shared context or inherent connections
created by additional hierarchical structures may further introduce
dependencies between stimuli, a statistical phenomenon commonly known as
clustering \citep{Everitt_et_al_2010}. Although the CJ literature
acknowledges the presence of such hierarchical structures, the
statistical handling of this extra source of dependency between stimuli
has been inadequate. For example, when CJ data includes multiple samples
of stimuli from the same individuals, researchers often rely on
(average) estimated BTL scores to conduct subsequent analyses and tests
at the individual hierarchical level
\citep{Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.
However, this approach can introduce additional statistical and
measurement issues, which we discuss in Section~\ref{sec-theory-issue2}.

In any case, the psychometric and statistical literature emphasizes the
need to address factors that create dependencies between stimuli, as
failing to do so can affect the reliability of the comparison outcomes
and lead to inaccurate conclusions about the trait's estimates. For
instance, researchers who overlook additional relevant traits, such as
judges' biases, can cause dimensional mismatches in the statistical
model used for analysis. This mismatch can artificially inflate the
reliability of the trait \citep[pp.~341]{Hoyle_et_al_2023} or, even
worse, introduce bias into the trait's estimates \citep{Ackerman_1989}.
Furthermore, as discussed in Section~\ref{sec-theory-issue1a},
researchers who exclude judges based on misfit statistics can risk
discarding valuable information, further biasing the trait's estimates
\citep[chap.~12]{Zimmerman_1994, McElreath_2020}. Lastly, researchers
who fail to account for hierarchical (grouping) structures can reduce
the precision of model parameter estimates, which may amplify the
overestimation of the trait's reliability
\citep[pp.~482]{Hoyle_et_al_2023}.

@@

In aggregate, researchers' inadequate handling of violations to the
assumptions of equal dispersion and zero correlation between stimuli,
along with the apparent disconnect between CJ's approach to trait
measurement and hypothesis testing, can undermine the reliability of the
trait and ultimately compromise its validity
\citep[pp.~2]{Perron_et_al_2015}. Consequently, adopting a more
systematic and integrated approach to examining what happens when judges
compare two stimuli could offer several statistical and measurement
benefits, including addressing these issues.

@@

This section presents a theoretical model for CJ that extends
Thurstone's theory. The model systematically incorporates all factors
involved when judges make pairwise comparisons. Additionally, the
section develops the statistical translation of the theoretical model
based on assumptions informed by the CJ theory.
